---
title: "Gamestats Data Analysis"
author: "Gjeltema"
date: "July 20, 2015"
output: html_document
---

## Question on Game Statistics  
Visualize the following data and ... 

## Importing the data
```{r getData}
suppressPackageStartupMessages(library("dplyr"))
suppressPackageStartupMessages(library("ggplot2"))

# ---- get data ----
gamestats <- read.csv("Q3_data.csv")
str(gamestats)

# choose proper variable types
gamestats <- gamestats %>% 
  mutate(Player1ID = as.factor(Player1ID),
         Player2ID = as.factor(Player2ID));
# POSIX does not work well with 'mutate'
gamestats$DateTime <- as.POSIXct(as.character(gamestats$DateTime), 
                                 format="%m/%d/%Y %H:%M");
glimpse(gamestats) # check formats
```

## Visually exploring the data
```{r plotSCoreSummary}
# ---- tidy the data (one observation per row)
#      by giving each player one single row    ----
gamestats.combo <- 
  # put player 1 into player 2
  gamestats %>% mutate(Player2ID = Player1ID, Player2Score = Player1Score) %>%
  # combine this swapped data set with original data set
  rbind(., gamestats) %>% 
  # keep what is needed
  select(DateTime, PlayerID=Player2ID, PlayerScore=Player2Score)

# ---- boxplot of score distribution by player ----
gamestats.combo %>% 
  ggplot() + geom_boxplot(aes(x=PlayerID, y=PlayerScore, color=PlayerID)) +
  labs(title="Distribution of Scores by Player Across All Games") + 
  theme(legend.position="none")

# clean up
rm(gamestats.combo)
```
We see that several players have played only one game or had the same score in each game. 
We see a few outliers - very high points.

```{r plotOverTime}
# ---- plot all stats over time ----
gamestats %>%
  ggplot(aes(x=DateTime)) + ylab("Player Score") +
  labs(title="Games Played Over Time", color ='Player ID') +
  geom_point(aes(y=Player1Score, color=Player1ID)) + 
  geom_point(aes(y=Player2Score, color=Player2ID)) + 
  guides(color=guide_legend(ncol=2)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1));
```
We see 3 distinct groups of events. Let us separate them and inspect them individually. 

## Segmenting the data
```{r segmentEvents}
# time and date separates the three distinct gaming groups
cutoff_date1 <- as.POSIXct("2015-05-27 02:00:01");
cutoff_date2 <- as.POSIXct("2015-05-28 00:00:01");

# ---- for each of the 3 distinct groups of activities over time 
#      we track the number of wins of each player                ----
gamer.group1 <- gamestats %>% filter(DateTime < cutoff_date1) %>%
  mutate(Wins.Player1 = cumsum(Player1Score > Player2Score),
         Wins.Player2 = cumsum(Player1Score < Player2Score),
         Ties = cumsum(Player1Score == Player2Score)) %>%
  droplevels();

gamer.group2 <- gamestats %>% filter((DateTime > cutoff_date1) &
                                     (DateTime < cutoff_date2)) %>%
  mutate(Wins.Player1 = NA, # not applicable, player ID changes
         Wins.Player2 = cumsum(Player1Score < Player2Score),
         Ties = cumsum(Player1Score == Player2Score)) %>%
  droplevels(); 

gamer.group3 <- gamestats %>% filter(DateTime > cutoff_date2) %>%
  mutate(Wins.Player1 = cumsum(Player1Score > Player2Score),
         Wins.Player2 = cumsum(Player1Score < Player2Score),
         Ties = cumsum(Player1Score == Player2Score)) %>%
  droplevels();

# ---- summary statistics on each group ----
gamer.group1 %>% summary
gamer.group2 %>% summary
gamer.group3 %>% summary
```

## Basic plots of the 3 segments of data
```{r basicSegmentPlots}
# ---- plot all stats as facets in their own small time frame ----
# create distinct header for each of the 3 separate groups over time
gamestats.periods <- gamestats %>% 
  mutate(TimePeriod = ifelse(DateTime < cutoff_date1, 
                             paste("1.) Morning of",as.Date(cutoff_date1)), 
                             ifelse(DateTime > cutoff_date2, 
                                    paste("3.) Afternoon of", as.Date(cutoff_date2)),
                                    paste("2.) Afternoon of", as.Date(cutoff_date1)))));

gamestats.periods %>%
  ggplot(aes(x=DateTime)) + 
  facet_wrap(~TimePeriod, scale="free_x") +
  scale_y_continuous(limits = c(0, NA)) +
  geom_line(aes(y=Player1Score, color=Player1ID, group=Player1ID)) + 
  geom_line(aes(y=Player2Score, color=Player2ID, group=Player2ID)) +
  geom_point(aes(y=Player1Score, color=Player1ID, group=Player1ID)) + 
  geom_point(aes(y=Player2Score, color=Player2ID, group=Player2ID)) +
  labs(title="Scores In Three Small Time Segments", y="Player Score", color ='Player ID') +
  guides(color=guide_legend(ncol=2))

# ---- plot each group individually with info on who is leading ----
gamer.group1 %>%
  ggplot(aes(x=DateTime)) +
  labs(title="Segment 1", y="Player Score", color ='Player ID') +
  geom_line(aes(y=Player1Score, color=paste("Player1, ID",Player1ID))) + 
  geom_line(aes(y=Player2Score, color=paste("Player2, ID",Player2ID))) +
  geom_point(aes(y=Player1Score, color=paste("Player1, ID",Player1ID))) + 
  geom_point(aes(y=Player2Score, color=paste("Player2, ID",Player2ID))) +
  geom_text(aes(y=0, label=paste0(Wins.Player1,":",Wins.Player2)), color=grey);

gamer.group2 %>%
  ggplot(aes(x=DateTime)) + 
  labs(title="Segment 2: (changing) Player 1  vs. (constant) Player 2", 
       y="Player Score", color='Player ID') +
  geom_point(aes(y=Player1Score, color="Player1")) + 
  geom_point(aes(y=Player2Score, color="Player2")) + 
  facet_wrap(Player1ID ~ Player2ID)#, scales="free")

gamer.group2 %>%
  ggplot(aes(x=DateTime)) + 
  geom_point(aes(y=Player1Score, color=Player1ID)) + 
  geom_point(aes(y=Player2Score)) +
  geom_line(aes(y=Player2Score)) +
  geom_text(aes(y=0, label=paste0("_:",Wins.Player2))) + # wins of player 2
  annotate("text", size=4,
           x=gamer.group2$DateTime[10],
           y=gamer.group2$Player2Score[10]*1.5, 
           label=paste("Player2 ID", gamer.group2$Player2ID[10])) +
  labs(title="Segment 2: Several (changing) Player 1 vs. One (consistent) Player 2", 
       y="Player Score", color='Player1 ID') 

gamer.group3 %>%
  ggplot(aes(x=DateTime)) + 
  labs(title="Segment 3", y="Player Score", color='Player ID') + 
  geom_line(aes(y=Player1Score, color=paste("Player1, ID",Player1ID))) + 
  geom_line(aes(y=Player2Score, color=paste("Player2, ID",Player2ID)))  +
  geom_text(aes(y=0, label=paste0(Wins.Player1,":",Wins.Player2)), color=grey);

rm(cutoff_date1, cutoff_date2)
```

## Advanced plots of the 3 segments  
We draw a plot with scores over time at the center (bottom left), number of wins 
of each player at the top left, legend aka player IDs at the top right, and point 
distribution by player on the bottom right side.  
```{r advancedPlotFct, echo=TRUE}
suppressPackageStartupMessages(library("gtable"))
suppressPackageStartupMessages(library("gridExtra"))

# ---- gets only the legend from a plot with legend ----
#' @description    function for extracting a legend from a plot
#' @input a.gplot  a ggplot2 object with legend
#' @output         a gtable object containing only the legend of the ggplot2 object
#' @example        plegend.group1 <- g_legend(center_plot(gamer.group1))
#'
#' code from https://github.com/hadley/ggplot2/wiki/Share-a-legend-between-two-ggplot2-graphs 
#' and http://stackoverflow.com/questions/13649473/add-a-common-legend-for-combined-ggplots
g_legend <- function(a.gplot){
  tmp <- ggplot_gtable(ggplot_build(a.gplot))
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  legend <- tmp$grobs[[leg]]
  return(legend)
  }


# ---- main plot with legend; 
#      returns a ggplot object incl. legend ----
center_plot<- function(data_set) {
  # extract the date part of date&time for the x-axis label
  x_axis_label <- paste("Time on", as.Date(data_set$DateTime[1]))
  
  pcenter.group <- data_set %>%
    ggplot(aes(x=DateTime)) + 
    scale_y_continuous(limits = c(0, NA)) +
    geom_point(aes(y=Player1Score, color=Player1ID)) + 
    geom_point(aes(y=Player2Score, color=Player2ID)) +  
    geom_line(aes(y=Player2Score, group=Player2ID), colour="gray75", size=0.5) +
    geom_line(aes(y=Player1Score, group=Player1ID), colour="gray75", size=0.5) + 
    theme(plot.margin = unit(c(0.7, 0.7, 0.5, 0.5), "lines")) +
    labs(x=x_axis_label, y="Player Score", color='Player ID') + 
    guides(color=guide_legend(ncol=1));
  
  return(pcenter.group)
  }

# ---- step function with number of wins ----
# only plots the wins of player 2, used only for 2nd segment
top_left_plot_onlyPlayer2 <- function(data_set) {
  ptop.group <- data_set %>%
    ggplot(aes(x=DateTime)) +
    # step function for player 2
    geom_step(aes(y=Wins.Player2, color=Player2ID)) +
    # adjust for y-axis to include 0
    scale_y_continuous(limits = c(-0.1, NA)) +
    theme(legend.position="none",
          axis.text.x = element_blank(), 
          axis.ticks.x = element_blank(),
          plot.margin = unit(c(1, 0.7, -0.5, 0.5), "lines")) + 
    labs(x=NULL, y=paste("Games Won \n by P.", 
                                      gamer.group2$Player2ID[1]))
  
  return(ptop.group)
  }

# plots the wins of players 1 & 2, used for 1st and 3rd segment
top_left_plot <- function(data_set) {
  ptop.group <- 
    top_left_plot_onlyPlayer2(data_set) +
    # include step function for player 1 as well
    geom_step(data=data_set, aes(x=DateTime, y=Wins.Player1, color=Player1ID)) +  
    # set y-label to general description
    labs(y="Games Won \n ");
  
  return(ptop.group)
  }


# ---- score density distribution  ----
bottom_right_plot <- function(data_set) {
  pright.group <- data_set %>%
    ggplot() + 
    geom_line(aes(x=Player1Score, color=Player1ID), stat="density") + 
    geom_line(aes(x=Player2Score, color=Player2ID), stat="density") + 
    scale_x_continuous(limits = c(0, NA)) +
    theme(legend.position="none", 
          axis.text.y = element_blank(),
          axis.ticks = element_blank(), 
          plot.margin = unit(c(0.7, 1, 0.5, -0.5), "lines")) + 
    labs(x=NULL, y="Score Density") +
    coord_flip()
  
  return(pright.group)
  }
```

### Segment 1 overall plot
```{r plotSeg1}
grid.arrange(
  arrangeGrob(top_left_plot(gamer.group1), 
              g_legend(center_plot(gamer.group1)), 
              center_plot(gamer.group1) + theme(legend.position="none"), 
              bottom_right_plot(gamer.group1), 
              widths=c(8,2), heights=c(3,7))
  )
```

### Segment 2 overall plot
```{r plotSeg2}
grid.arrange(
  arrangeGrob(top_left_plot_onlyPlayer2(gamer.group2),               
              g_legend(center_plot(gamer.group2) + # 2-column legend
                         guides(color=guide_legend(ncol=2))),              
              center_plot(gamer.group2) + theme(legend.position="none"), 
              bottom_right_plot(gamer.group2),               
              widths=c(8,2), heights=c(4,7))
  )
```

### Segment 3 overall plot
```{r plotSeg3}
grid.arrange(
  arrangeGrob(top_left_plot(gamer.group3), 
              g_legend(center_plot(gamer.group3)), 
              center_plot(gamer.group3) + theme(legend.position="none"), 
              bottom_right_plot(gamer.group3), 
              widths=c(8,2), heights=c(3,7))
  )
```
