The Long History Of Musicians Telling Republicans To Stop Playing Their Music  
Data From FiveThirtyEight  
As A Visualization  
Gjeltema, 6/19/2015  
========================================================  
The FiveThirtyEight blog had an article about which singers did not allow 
political candidates to use their songs 
(see http://fivethirtyeight.com/datalab/the-long-history-of-musicians-telling-republicans-to-stop-playing-their-music/
). There was no graphic in the article, the data was written as numbered points, 
which made me think about a visual representation ...  

# Import data
```{r dataMusiciansCandidates}
suppressPackageStartupMessages(library("stringr"))
suppressPackageStartupMessages(library("dplyr"))
suppressPackageStartupMessages(library("ggplot2"))
suppressPackageStartupMessages(library("grid"))
                               
# the manual way is faster for this data set,
# but it is interesting to try doing this purely in R.

# copy the data from website 
data_from_FiveThirtyEight <- c(
  "    Springsteen objected to Reagan’s use of the song “Born in the U.S.A.” during the 1984 election.", 
  "    Reagan also got dinged in 1984 by John Cougar Mellencamp for “Pink Houses.”",
  "    Bobby McFerrin objected to George H.W. Bush using the song “Don’t Worry, Be Happy” in 1988.",
  "    Sam & Dave objected to Bob Dole using the song “Soul Man” in 1996.",
  "    Springsteen objected to Dole using “Born in the U.S.A.” in 1996.",
  "    Mellencamp didn’t grant any more leeway to George W. Bush than to Reagan. Mellencamp complained when Bush used “R.O.C.K. in the U.S.A.” in 2000.",
  "    Tom Petty objected to Bush using the song “I Won’t Back Down” in 2000.",
  "    That same election, Sting objected to Bush using “Brand New Day.”",
  "    In a rare bit of bipartisanship in 2000, Sting also objected to Al Gore using “Brand New Day.”",
  "    John Hall of the band Orleans objected to Bush using the song “Still the One” during his re-election campaign in 2004.",
  "    Boston objected to Mike Huckabee using “More Than a Feeling” in 2008.",
  "    Van Halen objected to John McCain using “Right Now” in 2008.",
  "    Mellencamp is back! He complained about McCain using “Our Country” in 2008.",
  "    In the same election, Mellencamp had a problem with McCain using the song “Pink Houses.”",
  "    Heart objected to McCain using “Barracuda.”",
  "    Jackson Browne objected to McCain using “Running on Empty.”",
  "    Bon Jovi objected to McCain using “Who Says You Can’t Go Home.”",
  "    The Foo Fighters stopped McCain from playing “My Hero.”",
  "    Tom Petty objected to McCain using “I Won’t Back Down.”",
  "    ABBA complained about McCain using “Take a Chance on Me.”",
  "    Sam Moore of Sam & Dave objected to Barack Obama using “Hold On, I’m Coming” during the 2008 election.",
  "    Survivor objected to Newt Gingrich using the song “Eye of the Tiger” in 2012.",
  "    Gingrich also was dinged by The Heavy for “How You Like Me Now.”",
  "    Also in 2012, Dee Snider of Twisted Sister objected to Mitt Romney using “We’re Not Gonna Take It.”",
  "    Silversun Pickups stopped Romney from playing “Panic Switch.”",
  "    K’Naan objected to Romney using the song “Wavin’ Flag.”",
  "    Survivor objected to Romney using “Eye of the Tiger.”",
  "    Tom Petty complained about Michele Bachmann using “American Girl” in 2012.",
  "    Katrina and the Waves objected to Bachmann playing “Walking on Sunshine.”",
  "    And finally, now we have Young’s complaint about Trump’s use of “Rockin’ in the Free World.”"
  )


data_from_FiveThirtyEight <- 
  data_from_FiveThirtyEight %>% 
  str_replace_all("\\.",""); # get rid of all dots

# presidental campaign year  ====
usage_year <- data_from_FiveThirtyEight %>% str_extract("[1-2][0-9]{3}") # some NA

# song titles  ====
song_name <- data_from_FiveThirtyEight %>% 
  str_extract("“.+”") %>%
  str_replace(pattern="“", replacement="") %>% 
  str_replace(pattern="”", replacement="");
  
# names (of politicians and musicians) are capitalized  ====
nouns <- data_from_FiveThirtyEight %>%
  # remove song titles 
  str_replace(pattern=song_name, replacement="") %>%
  # separate words
  str_split(" ") %>%
  # extract words that are capitalized
  lapply(function(x) str_extract(x, "[A-Z].+")) %>%
  # merge individual words back to sentence, leaving only capitalized names
  sapply(function(x) paste(x[!is.na(x)], collapse=" ")) %>%
  # get rid of 's string
  str_replace_all(pattern="’s", replacement="") %>%
  # get rid of stopwords "And", "Also", "That", and "In"  
  str_replace_all(pattern="And ", replacement="") %>%
  str_replace_all(pattern="Also ", replacement="") %>%
  str_replace_all(pattern="That ", replacement="") %>%
  str_replace_all(pattern="In ", replacement="");

# get politician  ====
# seems like last word is usually the candidate's name --> parse it out
politician <- word(nouns,-1) 
# there seem to be exceptions to 'musician before candidate'
last_word_not_politic <- c(2,23)
politician[last_word_not_politic] <- word(nouns[last_word_not_politic], 1)
# account for 2 people with same last name
politician[3] <- paste("G.H.W.", politician[3]) # first substitute father
politician[politician == "Bush"] <- "G.W. Bush" # then substitute son
rm(last_word_not_politic);

# get musician ====
musician <- 
  nouns %>% 
  str_replace_all(pattern=politician, replacement="") %>% 
  str_trim(); 
# need to do a few corrections 
musician[2]  <- word(musician[2], 3)
musician[4]  <- "Sam & Dave"
musician[21] <- musician[4]
musician[24] <- word(musician[24], 3, 4)
musician[29] <- "Katrina and the Waves"
musician[30] <- paste("Neil", musician[30])
first_word_only           <- c(6, 8, 9, 11, 13, 22)
musician[first_word_only] <- word(musician[first_word_only], 1)
first_two_words_only           <- c(3, 7, 10, 12, 28) 
musician[first_two_words_only] <- word(musician[first_two_words_only], 1, 2);
rm(first_word_only, first_two_words_only);

# get our data frame ====
data_songs <- data.frame(
  Musician   = musician %>% str_replace_all("’", "'"),
  Song       = song_name %>% str_replace_all("’", "'"),
  Politician = politician,
  Year       = as.numeric(usage_year)
  );

# fill in missing values where known ====
data_songs <- data_songs %>% 
  mutate(Year = ifelse((is.na(Year) & (Politician=="McCain")), 2008, Year),
         Year = ifelse((is.na(Year) & (Politician=="G.W. Bush")), 2000, Year),
         Year = ifelse((is.na(Year) & (Politician %in% c("Gingrich", "Romney", "Bachmann"))), 2012, Year),
         Year = ifelse((is.na(Year) & (Politician=="Trump")), 2015, Year)
         );

# stats ====
data_songs
summary(data_songs)
```

# Create visualizations  
Now that the data is in good shape, let us create some plots in the 
Fivethirtyeight style (I'll be using the template from Austin Clemens,
http://austinclemens.com/blog/2014/07/03/fivethirtyeight-com-style-graphs-in-ggplot2/)  

## Politician vs. Musician
```{r dataVizByMusician, fig.width=10}
# ==== sort politician and musician by year and politician ====
data_songs5 <- data_songs %>% 
  mutate(Politician = factor(Politician, 
                             levels=unique(Politician[order(Year, Politician)]), 
                             ordered=TRUE),
         Musician = factor(Musician, 
                           levels=unique(Musician[order(Year, Politician, Musician)]), 
                           ordered=TRUE));

# ==== take care of overlapping songs from same musician (forbidden for
# same polititian in same year) by merging the two songs into 1 observation ====
data_songs5a <- data_songs5 %>% 
  filter(!(Politician=="McCain" & Musician=="Mellencamp"));
data_songs5b <- data_songs5 %>% 
  filter(Politician=="McCain" & Musician=="Mellencamp") %>%
  mutate(Song = "Our Country \n Pink Houses") %>% .[1,];
data_songs5combined <- bind_rows(data_songs5a, data_songs5b)

# ==== 538-style plot ====
ggplot(data_songs5combined) +
  geom_text(aes(x=Politician, y=Musician, label=Song, color=factor(Year), size=0.6)) +
  theme_bw() +
  # Set the entire chart region to a light gray color
  theme(panel.background=element_rect(fill="#F0F0F0")) +
  theme(plot.background=element_rect(fill="#F0F0F0")) +
  theme(panel.border=element_rect(colour="#F0F0F0")) +
  # Format the grid
  theme(panel.grid.major=element_line(colour="#D0D0D0",size=.75)) +
  theme(axis.ticks=element_blank()) +
  # Dispose of the legend
  theme(legend.position="none") +
  # Set title and axis labels, and format these and tick marks
  ggtitle("Presidental candidates prohibited from playing song during their campaign") +
  theme(plot.title=element_text(face="bold",hjust=0,vjust=2,colour="#3C3C3C",size=14)) +
  ylab("Musician") + 
  xlab(paste0("Politician;   Campaign Year is colorcoded [",min(data_songs5combined$Year),
              "-",max(data_songs5combined$Year),"]")) +
  theme(axis.text.x=element_text(size=9,colour="#535353",face="bold")) +
  theme(axis.text.y=element_text(size=9,colour="#535353",face="bold")) +
  theme(axis.title.y=element_text(size=9,colour="#535353",face="bold",vjust=1.5)) +
  theme(axis.title.x=element_text(size=9,colour="#535353",face="bold",vjust=-.5)) +
  # Big bold line at y=0
  geom_hline(yintercept=0,size=1.2,colour="#535353") +
  # Plot margins
  theme(plot.margin = unit(c(1, 1, .5, .7), "cm"))
```

## Year vs. Song & Musician
```{r dataVizBySongMusician, fig.width=10}
# ==== sort song & musician by year and politician ====
data_songs2 <- data_songs %>% mutate(
  SongMusi  = paste0(Song, " (", Musician, ")"),
  SongMusi2 = factor(SongMusi, levels=unique(SongMusi[order(Year, Politician, Musician, SongMusi)]), ordered=TRUE) 
  );

# ==== jiggle the names a bit where needed ====
data_songs2$Year[data_songs2$Musician=="Sting" & data_songs2$Politician=="Gore"] <- 2000.9;
data_songs2$Year[data_songs2$Musician=="Sting" & data_songs2$Politician=="G.W. Bush"] <- 1998.1;
data_songs2$Year[data_songs2$Musician=="Survivor" & data_songs2$Politician=="Romney"] <- 2013.5;
data_songs2$Year[data_songs2$Musician=="Survivor" & data_songs2$Politician=="Gingrich"] <- 2010.5;


# ==== 538-style plot ====
ggplot(data_songs2, aes(x=Year, y=SongMusi2)) +  
  geom_text(aes(x=Year, y=SongMusi2, label=Politician, color=Politician, size=1)) +
  theme_bw() +
  # Set the entire chart region to a light gray color
  theme(panel.background=element_rect(fill="#F0F0F0")) +
  theme(plot.background=element_rect(fill="#F0F0F0")) +
  theme(panel.border=element_rect(colour="#F0F0F0")) +
  # Format the grid
  theme(panel.grid.major=element_line(colour="#D0D0D0",size=.75)) +
  scale_x_continuous(minor_breaks=0,breaks=c(1984, 1988,seq(1996,2012,4),2015),limits=c(1984,2015)) +
  scale_y_discrete() + 
  theme(axis.ticks=element_blank()) +
  # Dispose of the legend
  theme(legend.position="none") +
  # Set title and axis labels, and format these and tick marks
  ggtitle("Presidental candidates prohibited from playing song \n during their campaign") +
  theme(plot.title=element_text(face="bold",hjust=0,vjust=2,colour="#3C3C3C",size=15)) +
  ylab("Song (Musician)") + xlab("Campaign Year") +
  theme(axis.text.x=element_text(size=10,colour="#535353",face="bold")) +
  theme(axis.text.y=element_text(size=9,colour="#535353",face="bold")) +
  theme(axis.title.y=element_text(size=10,colour="#535353",face="bold",vjust=1.5)) +
  theme(axis.title.x=element_text(size=9,colour="#535353",face="bold",vjust=-.5)) +
  # Big bold line at y=0
  geom_hline(yintercept=0,size=1.2,colour="#535353") +
  # Plot margins and finally color
  #scale_colour_brewer(palette="Paired") +
  theme(plot.margin = unit(c(1, 1, .5, .7), "cm")) 
```